\part{Introduction}

\chapter{About this Document} 

\section{Who should read this Document}

This document is targetted at embedded software developers who want to get
started with mbeddr. It provides a hands-on introduction to all aspects of
mbeddr. The document does not (yet) discuss building custom extensions, it just
discusses how to work with the existing ones.


\section{Structure of this Document}

This document is structured into three major parts.

\parhead{Part I} starts with an introduction to the idea of and concepts
behind mbeddr in \sect{mbeddrintro}. We also discuss how mbeddr compares to
other embedded software development approaches, and discuss when to use mbeddr.
You may skip this section if you already get the idea of mbeddr.

\parhead{Part II} is the reference documentation. It discusses how to
install mbeddr (as well as MPS and all the other required tools), discusses
important keyboard shortcuts, walks you through a Hello World and discusses all
the default extensions of mbeddr incl. components, state machines, physical
units, exceptions, requirements tracing, product line variability and the formal
analyses. This part is indended to provide a solid overview over all existing
extensions.

\parhead{Part III} contains a comprehensive tutorial, which, based on the Hello
World from Part I, develops a non-trivial (but, of course, still small compared
to real-world software) program using various features of mbeddr. If you want to
understand what it feels like to use mbeddr, this is the part to work through.

\section{Feedback and Contact}

 