\section{Product Line Variability}

mbeddr supports two kinds of variability: runtime and static. \emph{Runtime
variability} makes the decision about which variant to run as the program
executes. The binary contains the code for all options. Since the code must be
aware of the variability, the underlying language must know about variability.
\emph{Static variability} makes the decision before program execution. In
particular, the MPS generators remove all the program code that is not part of a
particular variant. The binary is tailored, and the mechanism is generic --- the
target language does not have to be aware of the variability. We discuss both
approaches in this chapter.

Both approaches have in common that in mbeddr, you first describe the available
variability on an abstract level that is unrelated to the implementation
artifacts that realize the variability. We use feature models for this aspect. A
feature model describes the available variability (aka the configuration space)
in a software system. Let us start by creating a feature model that describes
variability for processing flights. To do so, add the
\ic{com.mbeddr.variability} devkit to your model and add a
\ic{VariabilitySupport} node into your program. After giving it a name, it looks
as follows:
 
\begin{lstlisting}[language=mbeddr, morekeywords={Variability, Support}]
Variability Support FlightVariability
\end{lstlisting}

In this node you can now create a feature model. It has a name, and we specify
the root feature \ic{processing}.

\begin{lstlisting}[language=mbeddr, morekeywords={Variability, Support,
feature, model}] 
feature model FlightProcessor { 
  processing
\end{lstlisting}

We now add two sub-features \ic{nullify} and \ic{normalizeSpeed}. You can add
children via an intention. By default, children are or-features (marked by the
\ic{?} in the parent), which means that each of the children can be in the
system or not.

\begin{lstlisting}[language=mbeddr, morekeywords={Variability, Support,
feature, model}] 
feature model FlightProcessor 
  processing ? { 
    nullify                                                    
    normalizeSpeed  
  }
\end{lstlisting}

Let us add children to \ic{normalizeSpeed}: \ic{maxCustom} and \ic{max100}. They
are xor, so only one of them can be in a configuration. \ic{maxCustom} has an
attribute \ic{maxSpeed}:

\begin{lstlisting}[language=mbeddr, morekeywords={Variability, Support,
feature, model}] 
feature model FlightProcessor 
  processing ? { 
    nullify                                                    
    normalizeSpeed xor { 
      maxCustom [int16/mps/ maxSpeed]
      max100
    }
  }
\end{lstlisting}

So here is what this all means: this feature model describes the variability of
a flight processor (as usual...). If \ic{nullify} is selected, the processor
sets the altitude to 0 --- we had seen this before. \ic{normalizeSpeed} changes
the speed. If the speed is over 100 (\ic{max100}), it is set to 100. If
\ic{maxCustom} is selected and the speed is over \ic{maxSpeed}, it is set to
\ic{maxSpeed}\footnote{I do realize that the example is getting more and more
hairbrained, but what the heck ... it does illustrate mbeddr :-)}. Notice how
this is purely conceptual variability, and we haven't connected it to
implementation code.

We can now go ahead and define configurations. These are instances of feature
models, i.e. some of the features are selected, other aren't. A configuration
has to be valid wrt. to its feature model: for example, a configuration cannot
have \ic{maxCustom} and \ic{max100} selected at the same time, since those
two features are mutually exclusive (\ic{xor}). Here is the simplest possible
configuration. It has no feature except the mandatory root feature:

\begin{lstlisting}[language=mbeddr, morekeywords={Variability, Support,
feature, configuration, model, configures}] 
configuration model cfgDoNothing configures FlightProcessor 
  processing { 
  }
\end{lstlisting}

Let us create another one by copying this one and then selecting features
(press \ic{Ctrl-Space} between \ic{processing}'s curlies):

\begin{lstlisting}[language=mbeddr, morekeywords={Variability, Support,
feature, configuration, model, configures}] 
configuration model cfgNullifyOnly configures FlightProcessor 
  processing { 
    nullify 
  }
\end{lstlisting}

A third configuration includes the \ic{maxCustom} feature. You can add the the
value of \ic{maxSpeed} with an intention. Note that if you tried to add
\ic{max100} as well you'd get an error --- the two are mutually exclusive.

\begin{lstlisting}[language=mbeddr, morekeywords={Variability, Support,
feature, configuration, model, configures}] 
configuration model cfgNullifyMaxAt200 configures FlightProcessor 
  processing { 
    nullify 
    normalizeSpeed { 
      maxCustom [maxSpeed = 200 mps] 
    } 
  } 
\end{lstlisting}



\subsection{Runtime Variability}

As mentioned above, runtime variability means that we will evaluate a
configuration \emph{at runtime} and based on the evaluation, make decisions
about program execution. To try this out, let us create a new module with the
beginnings of a test case in it. Make sure you call the test case from
\ic{Main}. 
 
\begin{lstlisting}[language=mbeddr, morekeywords={}] 
module RuntimeVariability imports FunctionPointers { 
   
  exported test case testRuntimeVar { 
    Trackpoint tp = { 
      id = 1 
      timestamp = 0 s 
      x = 0 m 
      y = 0 m 
      alt = 50 m 
      speed = 220 mps 
    };
  } 
}
\end{lstlisting}

Now let's add variability support. It involves several steps. First we have to
create a representation of the feature model in the program (we generate a
\ic{struct} that holds configurations). We do that by adding the following
construct:

\begin{lstlisting}[language=mbeddr, morekeywords={feature, model, @, runtime}] 
module RuntimeVariability imports FunctionPointers { 
   
  feature model @ runtime for FlightProcessor; 
   
  exported test case testRuntimeVar {...} 
\end{lstlisting}

We can now create a function that processes trackpoints depending on the
configuration. There are two noteworthy things. First, we pass values of type 
\ic{fmconfig<FlightProcessor>} to the function, representing configurations for
the \ic{FlightProcessor} feature model (\ic{cfgDoNothing}, \ic{cfgNullifyOnly},
and \ic{cfgNullifyMaxAt200} are valid values for this type). Second we use the
\ic{variant} statement to make parts of the procedural code depent on the set of
selected features:

\begin{lstlisting}[language=mbeddr, morekeywords={fmconfig, variant},
deletekeywords={result}] 
Trackpoint processTrackpoint(fmconfig<FlightProcessor> cfg, Trackpoint tp) { 
  Trackpoint result; 
  variant<cfg> { 
    case (nullify && maxCustom) { 
      result = process_nullifyAlt(tp); 
      if (tp.speed > maxCustom.maxSpeed) { 
        result.speed = maxCustom.maxSpeed; 
      } 
    } 
    case (nullify && max100) { 
      result = process_nullifyAlt(tp); 
      if (tp.speed > 100 mps) { 
        result.speed = 100 mps; 
      }  
    } 
    case (nullify) { result = process_nullifyAlt(tp); } 
    default { result = process_doNothing(tp); } 
  } 
  return result; 
} 
\end{lstlisting}

Note that the \ic{case}s are evaluated top-to-bottom, so more specialized cases
must be higher in the list. Also, only one \ic{case} will ever be executed ---
no \ic{break} needed! Notice how the \ic{variant} construct is a new statement,
so this does depend on C. Currently we support only this statement, but
expressions or feature-dependent states (in state machines) would also be
feasible, of course.

We can now go back to our test case and write assertions, calling the
\ic{processTrackpoint} function with several configuration models. Below is the
first one. We first create a variable of type \ic{fmconfig<FlightProcessor>}
that holds a configuration (same type as in the argument to
\ic{processTrackpoint}). We then use the \ic{store config} construct to store a
configuration (\ic{cfgDoNothing}) into the \ic{cfg} variable\footnote{This
construct is a bit awkward, it would be nicer if this were an expression.
However, because of limitations in how C deals with structs this is not possible
without a performance hit. We didn't want to take that performance hit and have
opted to go with the uglier syntax.}. Then we call \ic{processTrackpoint} with
the configuration and the trackpoint and assert the result:

\begin{lstlisting}[language=mbeddr, morekeywords={fmconfig, variant, store,
config, into}, deletekeywords={result}] 
exported test case testRuntimeVar { 
  Trackpoint tp = {...}; 
   
  fmconfig<FlightProcessor> cfg; 
   
  store config<FlightProcessor, cfgDoNothing> into cfg; 
  Trackpoint res1 = processTrackpoint(cfg, tp); 
  assert(0) res1.alt == 50 m; 
  assert(1) res1.speed == 220 mps; 
}
\end{lstlisting}

We can store other configurations into \ic{cfg} and reuse the variable:

\begin{lstlisting}[language=mbeddr, morekeywords={fmconfig, variant, store,
config, into}, deletekeywords={result}] 
store config<FlightProcessor, cfgNullifyOnly> into cfg; 
Trackpoint res2 = processTrackpoint(cfg, tp); 
assert(2) res2.alt == 0 m; 
assert(3) res2.speed == 220 mps; 
 
store config<FlightProcessor, cfgNullifyMaxAt200> into cfg; 
Trackpoint res3 = processTrackpoint(cfg, tp); 
assert(2) res3.alt == 0 m; 
assert(3) res3.speed == 200 mps;
\end{lstlisting}


\subsection{Static Variability}


   
   




















