\part{Introduction}

\chapter{About this Document} 

\section{Who should read this Document}

This document is targetted at embedded software developers who want to get
started with mbeddr. It provides a hands-on introduction to all aspects of
mbeddr. The document does not (yet) discuss building custom extensions, it just
discusses how to work with the existing ones.


\section{Structure of this Document}

\parhead{Part I} starts with an introduction to the idea of and concepts
behind mbeddr in \sect{mbeddrintro}. We also discuss how mbeddr compares to
other embedded software development approaches, and discuss when to use mbeddr.
You may skip this section if you already get the idea of mbeddr.

\parhead{Part II} contains a comprehensive tutorial. It starts with a guide to
installing mbeddr, provides some tipps about using MPS and then runs through a
comprehensive tutorial of using mbeddr. It starts with a Hello World and then
adds elements from most of the mbeddr default extensions. If you are new to
mbeddr and want to understand what it feels like to use mbeddr, this is
the part to work through. This part also contains an set of extensive examples
or \emph{extending} mbeddr with new languages or C extensions.
 
\parhead{Part III} is the reference documentation. The part has a chapter
for each of the major features of mbeddr. In contrast to the tutorial it
tried so be complete. It also discusses the differences of mebddr C with
regular C. This part is indended to provide a solid reference for all existing
extensions. Use this part to look up details when you get stuck.


There is additional documentation you may want to take a look at. For example,
you can find a number of workshop, conference and journal papers that describe
mbeddr in more (academic) detail at \ic{http://mbeddr.wordpress.com/further-reading/} 
Also, there is a big and growing set of videos available at 
\ic{http://mbeddr.wordpress.com/screencasts/}.






\section{Feedback, Support and Contact}

While mbeddr is mature enough to be used for real-world software development, it
is still new and rough at the edges. To get support in case you run into
problems, or if you just share thoughts with the mbeddr developers, we have
create a Google Group:

\ \ \ \ic{https://groups.google.com/forum/?fromgroups\#!forum/mbeddr-discuss}

You can also submit issues at the github issue tracker:

\ \ \ \ic{https://github.com/mbeddr/mbeddr.core/issues}

















 